Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2018-03-15 paper-tourrd3.tex
submitted paper

The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. 2011) provides the computing and display
in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
of tours: grand, guided, little and local. The projection dimension can
be chosen between one and the number of variables in the data. The
display, though, has no capacity for interaction. The viewer can watch
the tour like a movie, but not pause it and restart, or change tour
type, or number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
choice today, and has been superceded with primarily web-capable tools,
like Shiny (Chang et al. 2017). To display dynamic graphics though, is
not straight-forward. This paper explains how to use D3 (Bostock,
Ogievetsky, and Heer 2011) as the display engine in a Shiny GUI, using
custom message passing between server and client.

@@

The \pkg{tourr} package (Wickham et al. 2011) is an R implementation of
the tour algorithms discussed in Cook et al. (2007). It includes methods
for geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example:

\begin{Schunk}
\begin{Sinput}
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device cannot be used
animate_dist(flea[, 1:6], center = FALSE)
\end{Sinput}
\end{Schunk}

generating the animation in Figure \ref{tour}.

\begin{figure}[ht]
EMBED MOVIE HERE
\end{figure}

@@

\subsection{Connecting the tour projections to D3 display using
\texttt{sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}

D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer 2011) is a
JavaScript library for manipulating documents based on data. The
advantages of D3 are similar to those provided by Shiny: namely, an
industry standard with rich array of powerful, easy to use methods and
widgets that can be displayed on a wide variety of devices, with a large
user base. D3 works on data objects in the JavaScript Object Notation
(JSON) format, which are then parsed and used to display customisable
data visualisations.

The new implementation of the tour interface uses D3 to render each
projection step returned by R, focusing on 2D projections as a test
case. It does this by drawing and re-drawing a scatterplot with dots (or
circles in D3 language) and providing SVG objects for the web browser to
render.

@@

The D3 canvas makes for smooth drawing and re-drawing of the data
projections. Adding a graphical user interface (GUI) around the display
is straightforward with the Shiny package, e.g.~control elements such as
stop/go, increase/decrease speed, change tour type, add/remove variables
from the mix.

@@

@@

The code is available at {[}\url{https://github.com/makipp/}
TourrGuiD3{]}, and the source material for this paper is available at
{[}\url{https://github.com/dicook/paper-tourrd3}{]}.

The \texttt{sendCustomMessage} function provides a way to share a tour
path with the D3 renderer, and embed it in a Shiny GUI providing
controls such as stop/go, increase/decrease speed, change tour type,
add/remove variables. However, the approach can result in inconsistent
speed which appears as jerkiness in the motion, and it isn't suitable
for large numbers of observations.

@@

\hyperdef{}{refs}{}

Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
GWidgets GUI for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

Lawrence, Michael, and Duncan {Temple Lang}. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

>>>>

The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
provides the computing and display in R (R Core Team
\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
tours: grand, guided, little and local. The projection dimension can be
chosen between one and the number of variables in the data. The display,
though, has no capacity for interaction. The viewer can watch the tour
like a movie, but not pause it and restart, or change tour type, or
number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham \protect\hyperlink{ref-tourrGui}{2012}), with
was programmed with the RGtk2 package (Lawrence and Temple Lang
\protect\hyperlink{ref-RGtk2}{2010}). This is not the toolkit of choice
today, and has been superceded with primarily web-capable tools, like
Shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
dynamic graphics though, is not straight-forward. This paper explains
how to use D3 (Bostock, Ogievetsky, and Heer
\protect\hyperlink{ref-D3}{2011}) as the display engine in a Shiny
graphical user interface (GUI), using custom message passing between
server and client.

@@

The \pkg{tourr} package (Wickham et al.
\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
algorithms discussed in Cook et al.
(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example:

\begin{Schunk}
\begin{Sinput}
library(tourr)
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
\end{Sinput}
\end{Schunk}

generating the animation, with some stills shown in Figure \ref{tour}.

\begin{figure}[ht]
\centerline{\includegraphics[width=5cm]{figures/tour1.png}\includegraphics[width=5cm]{figures/tour2.png}\includegraphics[width=5cm]{figures/tour3.png}}
\caption{Three projections from a 1D tour of 6D data, displayed as a density. Full video can be seen at https://vimeo.com/255466661.}
\label{tour}
\end{figure}

@@

\subsection{\texorpdfstring{Connecting the tour projections to D3
display using
\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}

D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer
\protect\hyperlink{ref-D3}{2011}) is a JavaScript library for
manipulating documents based on data. The advantages of D3 are similar
to those provided by Shiny: namely, an industry standard with rich array
of powerful, easy to use methods and widgets that can be displayed on a
wide variety of devices, with a large user base. D3 works on data
objects in the JavaScript Object Notation (JSON) format, which are then
parsed and used to display customisable data visualisations.

The new implementation of the tour interface uses D3 to render each
projection step returned by R, focusing on 2D projections as a test
case. It does this by drawing and re-drawing a scatterplot with dots (or
circles in D3 language) and providing SVG objects for the web browser to
render. Figure \ref{tourrD3} shows the new GUI.

\begin{figure*}[ht]
\centerline{\includegraphics[width=15cm]{figures/TourrD3.png}}
\caption{Shiny GUI for the tour, with D3 as the display engine. GUI provides controls to select tour type, change speed, restart, and select variables to include.}
\label{tourrD3}
\end{figure*}

@@

The D3 canvas makes for smooth drawing and re-drawing of the data
projections. Adding a GUI around the display is straightforward with the
Shiny package, e.g.~control elements such as stop/go, increase/decrease
speed, change tour type, add/remove variables from the mix.

@@

Another disadvantage is that the displays needs to be coded anew. D3
provides mostly primitives, and example code, to make scatterplots, and
contours, but the data displays all need to be coded again.

The custom message tools from Shiny provide a way to share a tour path
with the D3 renderer, and embed it in a Shiny GUI providing controls
such as stop/go, increase/decrease speed, change tour type, add/remove
variables. However, the approach doesn't provide the smooth motion that
is needed for easy display of projections, and is slow for large numbers
of observations.

@@

The code is available at \url{https://github.com/makipp/TourrGuiD3}, and
the source material for this paper is available at
\url{https://github.com/dicook/paper-tourrd3}.

\subsection{Acknowledgements}\label{acknowledgements}

Thanks to Yihui Xie for pointing out the custom message tools.

@@

\hypertarget{refs}{}
\hypertarget{ref-D3}{}
Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

\hypertarget{ref-shiny}{}
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\hypertarget{ref-gt_pp}{}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

\hypertarget{ref-gt_pp_mc}{}
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

\hypertarget{ref-tourrGui}{}
Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
gWidgets Gui for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

\hypertarget{ref-ihaka:1996}{}
Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

\hypertarget{ref-RGtk2}{}
Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

\hypertarget{ref-R}{}
R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\hypertarget{ref-tourr}{}
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

### 2018-03-15 submission/paper-tourrd3.Rmd
submitted paper

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine in a Shiny GUI, using custom message passing between server and client.

@@

```{r eval=FALSE, echo=TRUE}
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device cannot be used
animate_dist(flea[, 1:6], center = FALSE)
```

generating the animation in Figure \ref{tour}.

\begin{figure}[ht]
EMBED MOVIE HERE
\end{figure}

@@

The new implementation of the tour interface uses D3 to render each projection step returned by R, focusing on 2D projections as a test case. It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.

@@

The D3 canvas makes for smooth drawing and re-drawing of the data projections. Adding a graphical user interface (GUI) around the display is straightforward with the Shiny package, e.g. control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.

@@

The code is available at [https://github.com/makipp/ TourrGuiD3], and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3].

The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a Shiny GUI providing controls such as stop/go, increase/decrease speed, change tour type, add/remove variables. However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.

>>>>

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine in a Shiny graphical user interface (GUI), using custom message passing between server and client.

@@

```{r eval=FALSE, echo=TRUE}
library(tourr)
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
```

generating the animation, with some stills shown in Figure \ref{tour}.

\begin{figure}[ht]
\centerline{\includegraphics[width=5cm]{figures/tour1.png}\includegraphics[width=5cm]{figures/tour2.png}\includegraphics[width=5cm]{figures/tour3.png}}
\caption{Three projections from a 1D tour of 6D data, displayed as a density. Full video can be seen at https://vimeo.com/255466661.}
\label{tour}
\end{figure}

@@

The new implementation of the tour interface uses D3 to render each projection step returned by R, focusing on 2D projections as a test case. It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render. Figure \ref{tourrD3} shows the new GUI.

\begin{figure*}[ht]
\centerline{\includegraphics[width=15cm]{figures/TourrD3.png}}
\caption{Shiny GUI for the tour, with D3 as the display engine. GUI provides controls to select tour type, change speed, restart, and select variables to include.}
\label{tourrD3}
\end{figure*}

@@

The D3 canvas makes for smooth drawing and re-drawing of the data projections. Adding a GUI around the display is straightforward with the Shiny package, e.g. control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.

@@

The code is available at [https://github.com/makipp/TourrGuiD3](https://github.com/makipp/TourrGuiD3), and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3](https://github.com/dicook/paper-tourrd3).

## Acknowledgements

Thanks to Yihui Xie for pointing out the custom message tools.
